---
title: "Statistical testing of quantitative omics data"
author: "Veit Schwämmle"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{PolySTest}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

Welcome to the vignette of the PolySTest package. We will run statistical tests for differential abundances in a proteomics data set. 

## Installation

To install PolySTest, you can use Bioconductor's BiocManager:

```{r, eval=FALSE}
if (!requireNamespace("BiocManager", quietly = TRUE))
    install.packages("BiocManager")

BiocManager::install("PolySTest")
```

## Loading data
First, we load the necessary package and example data and reduce it to 200
lines. The data set consists of protein abundances in liver samples from mice
fed with four different diets. The data contains 3 replicates per diet.

```{r}
library(PolySTest)
library(SummarizedExperiment)

# Load the data file "LiverAllProteins.csv" from the PolySTest package
filename <- system.file("extdata", "LiverAllProteins.csv", package = "PolySTest")

dat <- read.csv(filename, row.names=1, stringsAsFactors = FALSE)
dat <- dat[1:200,] # Use a subset of the data for this example
```

## Creating a SummarizedExperiment object
We now create a `SummarizedExperiment` object from the data. This object will
contain the protein abundances and the sample metadata.

In addition, we normalize by the median of the columns to adjust for differences
in the sample amounts loaded into the instrument.

```{r}


sampleMetadata <- data.frame(Condition = rep(paste("Condition", 1:4), each=3),
                             Replicate = rep(1:3, each=4))

fulldata <- SummarizedExperiment(assays = list(quant = dat), colData = sampleMetadata)
rowData(fulldata) <- rownames(dat)
metadata(fulldata) <- list(NumReps = 3, NumCond = 4)

# Normalization (example)
dat <- t(t(dat) - colMedians(as.matrix(dat), na.rm = TRUE))
assay(fulldata, "quant") <- dat

fulldata <- update_conditions_with_lcs(fulldata)
```

## Defining comparisons
We define the comparisons we want to test. In this case, we want to compare
each condition with the first one. For that, we create a matrix with the
comparisons, where the first columnn contains the reference condition and the
second column contains the condition to be compared.

We can use the function `create_pairwise_comparisons` for this:


```{r}
conditions <- unique(colData(fulldata)$Condition)

allComps <- create_pairwise_comparisons(conditions, 1)
```

## Running the statistical tests
We can now run the statistical tests. We will use the `PolySTest_paired` function
to run paired tests. The function will add columns to the `rowData` of the
`SummarizedExperiment` object containing the p-values and FDRs for each comparison.

```{r}
fulldata_paired <- PolySTest_paired(fulldata, allComps)

```

## Running unpaired tests
We can also run unpaired tests using the `PolySTest_unpaired` function. The 
experimental design dictated whether to use paired or unpaired tests.

```{r}
fulldata_unpaired <- PolySTest_unpaired(fulldata, allComps)

```

## Visualization of test results
PolySTest provides several visualizations to compare the tests, the pairings, and 
number of significant hits for different FDR threshold.

# Plotting the results

While p-values need to be corrected for multiple testing to control the false
discovery rate (FDR), their distribution can provide insight into the 
test performances and the underlying data.

Here, select few comparison and tests as the number of panels will become quite
large, and the lead to the plot error "figure margins too large".

```{r}
# Define comparisons to visualize
comps <- allComps[1:2,]
# create _vs_ strings
comps <- paste0(comps[,2], "_vs_", comps[,1])

# Plotting the results
plotPvalueDistr(fulldata_unpaired,  comps, c("PolySTest", "limma", "Miss test"))
```

More common are volcano plots, here given for the same tests as above. We furthermore 
select the indicex of the proteins most significantly changing in the first comparison.

```{r}
# Select proteins with FDR < 0.01

sigProts <- which(rowData(fulldata_unpaired)[, paste0("FDR_PolySTest_", comps[1])] < 0.01)

# Volcano plot
plotVolcano(fulldata_unpaired, comps, sel_prots = sigProts, c("PolySTest", "limma", "Miss test"))
```

Now let's look into the overlap between the different tests. An upset plot
provides this information 

```{r}

```


## Session info
```{r}
sessionInfo()

```
